MATH 245 old and unused content

### 2024-02-17 PDEs/math245_hw2.Rmd
hw2 pde neumann

---
title: "MATH 245 Homework 2"
author: "Ruby Krasnow and Tommy Thach"
date: "2024-02-16"
output:
  pdf_document:
    extra_dependencies: ["derivative", "tcolorbox", "xcolor"]
---

@@

## Done, feel OK: 1, 2, 5, 7, 8
## Done, low confidence: 4
## Redo: 6
## Not started: 3

@@

Solve the Neumann boundary value problem for the wave equation on half line:
\begin{equation*}
    \begin{cases}
      u_{tt}=c^2u_{xx}+f(t,x) & 0<x<\infty\\
      u(0,x)=\phi{x} & 0<x<\infty\\
      u_t(0,x) = \psi{x} & 0<x<\infty\\
      u_x(t,0) = h(t) & t>0
    \end{cases}       
\end{equation*}

@@

Let's try the factorization method if $c=c(x)$:
$$\left( \partial_{t}+c(x)\partial_{x}\right) \left(\partial_{t}-c(x)\partial_{x}\right)u=0$$
Set $\xi=x+c(x)t$ and $\eta=x-c(x)t$.

By the chain rule,

$$\frac{\partial}{\partial x} = \frac{\partial}{\partial \xi} \frac{\partial \xi}{\partial x}+\frac{\partial}{\partial\eta}\frac{\partial\eta}{\partial x}$$

$$=\partial_{\xi}(1+c')+\partial_{\eta}\,(1-c')$$

$$\frac{\partial}{\partial t} = \frac{\partial}{\partial \xi} \frac{\partial \xi}{\partial t}+\frac{\partial }{\partial\eta}\frac{\partial\eta}{\partial t}$$
$$=c\partial_{\xi}-c\partial_{\eta}$$
When we plug these back into
$$\left( \partial_{t}+c\partial_{x}\right) \left(\partial_{t}-c\partial_{x}\right)u=0$$
we get

$$ \left(c\partial_{\xi}-c\partial_{\eta}+c\partial_{\xi}(1+c')+c\partial_{\eta}\,(1-c')\right)
\left(c\partial_{\xi}-c\partial_{\eta} -c\partial_{\xi}(1+c')-c\partial_{\eta}\,(1-c')\right)u=0$$

When $c$ is constant, $c'=0$ and this simplifies to $-4c^2 u_{\xi\eta}=0$, allowing us to integrate to find $u(x,t)=F(x+ct)+G(x-ct)$. But we cannot do the same simplification when we have the $c'$ terms, meaning we cannot solve the wave equation by the same method when the wave speed is not constant.

Similarly,

Let's try the factorization method if $c=c(t)$:
$$\left( \partial_{t}+c(t)\partial_{x}\right) \left(\partial_{t}-c(t)\partial_{x}\right)u=0$$
Set $\xi=x+c(t)t$ and $\eta=x-c(t)t$.

By the chain rule,

$$\frac{\partial}{\partial x} = \frac{\partial}{\partial \xi} \frac{\partial \xi}{\partial x}+\frac{\partial}{\partial\eta}\frac{\partial\eta}{\partial x}=
\partial_{\xi}+\partial_{\eta}$$

$$\frac{\partial}{\partial t} = \frac{\partial}{\partial \xi} \frac{\partial \xi}{\partial t}+\frac{\partial }{\partial\eta}\frac{\partial\eta}{\partial t}=\partial_{\xi}(c't+c)-\partial_{\eta}(c't+c)
$$

When we plug these back into
$$\left( \partial_{t}+c\partial_{x}\right) \left(\partial_{t}-c\partial_{x}\right)u=0$$
we get

$$\left( \partial_{\xi}(c't+c)-\partial_{\eta}(c't+c)+c\partial_{\xi}+c\partial_{\eta} \right)
\left(\partial_{\xi}(c't+c)-\partial_{\eta}(c't+c)-c\partial_{\xi}-c\partial_{\eta}\right)u=0$$
Where again we cannot cancel terms to simplify our equation as we could with constant $c$.

>>>>

---
title: "MATH 245 Homework 2"
author: "Ruby Krasnow and Tommy Thach"
date: "2024-02-17"
output:
  pdf_document:
    extra_dependencies: ["derivative", "tcolorbox", "xcolor"]
---

@@

@@

Solve the Neumann boundary value problem for the wave equation on half line:
\begin{equation*}
    \begin{cases}
      u_{tt}=c^2u_{xx}+f(t,x) & 0<x<\infty\\
      u(0,x)=\phi(x) & 0<x<\infty\\
      u_t(0,x) = \psi(x) & 0<x<\infty\\
      u_x(t,0) = h(t) & t>0
    \end{cases}       
\end{equation*}

The method we will use is to first convert to homogenous Neumann boundary conditions, and then extend the equation to the entire line using the reflection method.

First, define $w(t,x)=u(t,x)-H(t)$, where $H_x(t)=h(t)$.

Then our IBVP becomes:
\begin{equation*}
    \begin{cases}
      \square w= \square u - H''(t)=f(t,x)-H''(t) \equiv F(t,x) & x \geq 0, t>0\\
      w(0,x)=u(0,x)-H(0)=\phi(x) - H(0) \equiv \Phi(x) & x \geq 0\\
      w_t(0,x) =  u_t(0,x)  - H_t(0)= \psi(x) - H_t(0) \equiv \Psi(x) & x \geq 0\\
      w_x(0,x) = u_x(t,0) - H_x(0)= u_x(t,0)-h(t) \equiv 0 & t \geq 0
    \end{cases}       
\end{equation*}

Now, since we have homogeneous Neumann boundary conditions, we will use even extensions of the initial data to the whole line, namely:

\begin{equation} 
\Phi_{\text{even}}=
    \begin{cases} 
      \Phi(x) & \text{for } x \geq 0\\
      \Phi(-x) & \text{for } x \leq 0\\
    \end{cases}  
    ,\quad
\Psi_{\text{even}}=
    \begin{cases}
      \Psi(x) & \text{for } x \geq 0\\
      \Psi(-x) & \text{for } x \leq 0\\
    \end{cases}
    ,\quad
F_{\text{even}}=
    \begin{cases} 
      F(t,x) & \text{for } x \geq 0\\
      F(t,-x) & \text{for } x \leq 0\\
      \end{cases}
\end{equation}

Then we assume $U(t,x)$ is an extension of $w(t,x)$ to the whole line. Since the initial condition and non-homogeneous parts are both even, the solution $U$ is even, so we automatically have $U_x(t,0)=0$.

Consider
\begin{equation} 
    \begin{cases} 
      \square U(t,x)=F_{\text{even}}(t,x) & \text{for } x \in \mathbb{R}, t>0\\
      U(0,x)=\Phi_{\text{even}}(x) & \text{for } x \in \mathbb{R} \\
      U_t(0,x)=\Psi_{\text{even}}(x) & \text{for } x \in \mathbb{R}
    \end{cases} 
\end{equation}

The solution at point $(t_0, x_0)$ is

\begin{equation*}
U(t_0,x_0)= \frac{\Phi_{\text{even}}(x_0+ct_0)+\Phi_{\text{even}}(x_0-ct_0)}{2} +
\frac{1}{2c} \int_{x_0-ct_0}^{x_0+ct_0}\Psi_{\text{even}}(x)dx + \\
\frac{1}{2c} \int_{0}^{t_0}  \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} F_{\text{even}}(t,x) dx\,dt
\end{equation*}

Assume $t_0>0, x_0>0 \Rightarrow x_0+ct_0 >0$. Now assume $x_0>ct_0$, so the domain of dependence is entirely to the right of the line $x=ct$. This means $w(t_0, x_0)=U(t_0,x_0)$ and our solution is
\begin{equation*}
w(t_0,x_0)= \frac{\Phi(x_0+ct_0)+\Phi(x_0-ct_0)}{2} +
\frac{1}{2c} \int_{x_0-ct_0}^{x_0+ct_0}\Psi(x)dx + \\
\frac{1}{2c} \int_{0}^{t_0}  \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} F(t,x) dx\,dt 
\end{equation*}

\begin{equation*}
\frac{\phi(x_0+ct_0)-H(0)+\phi(x_0-ct_0)-H(0)}{2} +
\frac{1}{2c} \int_{x_0-ct_0}^{x_0+ct_0}\psi(x)-H'(0)dx + \\
\frac{1}{2c} \int_{0}^{t_0}  \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x)-H''(t) dx\,dt
\end{equation*}

\begin{align*}
& w(t_0,x_0)=\frac{\phi(x_0+ct_0)+\phi(x_0-ct_0)}{2}- H(0) +
\frac{1}{2c} \int_{x_0-ct_0}^{x_0+ct_0}\psi(x) dx - \frac{H'(0)}{2c}\left[(x_0+ct_0)-(x_0-ct_0)\right]+ \\
& \frac{1}{2c} \int_{0}^{t_0}  \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) dx\,dt -
\frac{1}{2c} H''(t) \int_{0}^{t_0}  \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} dx\,dt
\end{align*}

We can rewrite this as
\begin{align*}
& w(t_0,x_0)= \color{blue} \frac{\phi(x_0+ct_0)+\phi(x_0-ct_0)}{2} +
\frac{1}{2c} \int_{x_0-ct_0}^{x_0+ct_0}\psi(x) dx +
\frac{1}{2c} \int_{0}^{t_0}  \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) dx\,dt \\
\color{black}
& -H(0)-H'(0)t_0- \frac{1}{2c}\int_{0}^{t_0} H''(t) 2c(t_0-t)dt.
\end{align*}

Call the section in blue "A".

This means we have 
\begin{equation}\label{eqn:w1} 
w(t_0,x_0)=A-H(0)-H'(0)t_0- \int_{0}^{t_0} H''(t)(t_0-t)\,dt 
\end{equation}

Now we use integration by parts on the final integral in (\ref{eqn:w1}), setting
$$u=(t_0-t), \quad du=dt, \quad dv=H''(t)dt, \quad v=H'(t)$$,

so that 
$$\int_{0}^{t_0} H''(t)(t_0-t)\,dt =H'(t)(t_0-t)\vert_{0}^{t_0}+\int_{0}^{t_0} H'(t)dt=-H'(0)t_0+H(t_0)-H(0) $$
So (\ref{eqn:w1}) becomes

$$w(t_0,x_0)=A-H(0)-H'(0)t_0+H'(0)t_0-H(t_0)+H(0)$$
$$w(t_0,x_0)=A-H(t_0)$$

Since $u(t,x)=u(t,x)+H(t)$, we get $u(t_0, x_0)=A$, or
\begin{equation} \label{eqn:part1}
u(t_0, x_0)=\frac{\phi(x_0+ct_0)+\phi(x_0-ct_0)}{2} +
\frac{1}{2c} \int_{x_0-ct_0}^{x_0+ct_0}\psi(x) dx +
\frac{1}{2c} \int_{0}^{t_0}  \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) dx\,dt
\end{equation}

when $x_0>ct_0$.

Now we must consider what happens when $x_0<ct_0$, such that $x_0-ct_0<0$. Then we have
\begin{align*}
&w(t_0,x_0)= \frac{\phi(x_0+ct_0)-H(0)+\phi(ct_0-x_0)-H(0)}{2} \;+ \\
& \frac{1}{2c} \int_{x_0-ct_0}^{0} \psi(-x)-H'(0)\,dx\; + 
\frac{1}{2c} \int_{0}^{x_0+ct_0} \psi(x) -H'(0)\,dx \;+ \\
&\frac{1}{2c} \iint\limits_\Delta F_{\text{even}}(t,x) \, dx \, dt
\end{align*}

\begin{align*}
&w(t_0,x_0)= \frac{\phi(x_0+ct_0)+\phi(ct_0-x_0)}{2}-H(0) +\\
 & \color{violet} \frac{1}{2c} \int_{x_0-ct_0}^{0}\psi(-x)-H'(0)\,dx \;+ 
\frac{1}{2c} \int_{0}^{x_0+ct_0}\psi(x) -H'(0)\,dx \; \color{black} + \\
&\frac{1}{2c}\iint\limits_\Delta F_{\text{even}}(t,x)\,dx\,dt
\end{align*}

Violet section:
\begin{align*}
& \frac{1}{2c} \int_{x_0-ct_0}^{0}\psi(-x)\,dx \; + 
\frac{1}{2c} \int_{0}^{x_0+ct_0}\psi(x)\,dx \; 
-\frac{1}{2c} \left[ H'(0)(0-(x-ct_0))+H'(0)(x+ct_0-0) \right] = \\
& \frac{1}{2c} \left[\int_{0}^{ct_0-x_0}\psi(x)\,dx +  \int_{0}^{x_0+ct_0}\psi(x)\,dx \right]-H'(0)t_0
\end{align*}

Now we investigate $\frac{1}{2c}\iint\limits_\Delta F_{\text{even}}(t,x)\,dx\,dt$.

```{r, warning=FALSE, fig.height=3, fig.width=5, echo=FALSE}
c=1
t_0=5
x_0=3
x_1=seq(-20, 20, by=0.2)
x_2=seq(-20, 3, by=0.2)
x_3=seq(3, 20, by=0.2)
rm(x)

ggplot()+
  geom_line(aes(x=x_2, y=t_0-(1/c)*(x_0-x_2)))+
  geom_line(aes(x=x_3, y=t_0+(1/c)*(x_0-x_3)))+
  lims(y=c(0, 10), x=c(-5, 10))+
  geom_vline(xintercept=0, linetype="dashed")+
  geom_hline(yintercept = t_0-x_0/c, linetype="dashed")+
  geom_text(aes(x=-0.5, y=0.6, label="I"))+
  geom_text(aes(x=3, y=0.6, label="II"))+
  geom_text(aes(x=3, y=t_0-x_0/c+1, label="III"))+
  annotate("text", label="'('* x[0]* ','* t[0] *')'", x=3, y=5.5, size=3, parse = TRUE)+
  labs(x="x", y="t")+
  theme_light()
```

<!-- \newcommand{\Int}{\int\limits} -->

$$ \text{I}= \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{x_0-c(t_0-t)}^{0} F(t,-x)\;dx\;dt , \quad
\text{II}= \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{x_0+c(t_0-t)} F(t,x)\;dx\;dt,$$
$$\text{III}= \frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} F(t,x) \;dx\;dt $$

Now we will evaluate each integral separately:

$$ \text{I}= \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{x_0-c(t_0-t)}^{0} f(t,-x)-H''(t)\;dx\;dt$$
$$ = \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{x_0-c(t_0-t)}^{0} f(t,-x)\;dx\;dt -\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}}H''(t) \int_{x_0-c(t_0-t)}^{0}dx\;dt$$
$$ = -\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{-x_0+c(t_0-t)}^{0} f(t,x)\;dx\;dt +\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}}H''(t) (x_0-c(t_0-t))\;dt$$
$$ = \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{c(t_0-t)-x_0} f(t,x)\;dx\;dt +\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}}H''(t) (x_0-c(t_0-t))\;dt$$

$$\text{II}= \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{x_0+c(t_0-t)} f(t,x)\;dx\;dt-
\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} H''(t)\int_{0}^{x_0+c(t_0-t)}dx\;dt$$

$$= \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{x_0+c(t_0-t)} f(t,x)\;dx\;dt-
\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} H''(t)(x_0+c(t_0-t))\;dt$$

$$\text{III}= \frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) \;dx\;dt -\frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} H''(t) \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} dx\;dt $$
$$ = \frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) \;dx\;dt -\frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} H''(t) \;2c(t_0-t)\;dt $$
$$ = \frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) \;dx\;dt -\int_{t_0-\frac{x_0}{c}}^{t_0} H''(t) \;(t_0-t)\;dt $$
Now we use integration by parts on the final integral in the above expression, setting
$$u=(t_0-t), \quad du=dt, \quad dv=H''(t)dt, \quad v=H'(t)$$

so that 
$$\int_{t_0-\frac{x_0}{c}}^{t_0}  H''(t)(t_0-t)\,dt =H'(t)(t_0-t)\vert_{t_0-\frac{x_0}{c}}^{t_0}+\int_{t_0-\frac{x_0}{c}}^{t_0}  H'(t)dt$$
$$=-\frac{x_0}{c} H'\left(t_0-\frac{x_0}{c} \right)+H(t_0)-H\left(t_0-\frac{x_0}{c} \right)$$

So $\text{III}$ is
$$\frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) \;dx\;dt+\frac{x_0}{c} H'\left(t_0-\frac{x_0}{c} \right)-H(t_0)+H\left(t_0-\frac{x_0}{c} \right) $$

And now we can find the sum of the three regions:

\begin{align*}
\text{I}+\text{II}+\text{III}&= \frac{1}{2c} \int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{c(t_0-t)-x_0} f(t,x)\;dx\;dt +\frac{1}{2c} \int_{0}^{t_0-\frac{x_0}{c}} H''(t) (x_0-c(t_0-t))\;dt + \\
& \frac{1}{2c} \int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{x_0+c(t_0-t)} f(t,x)\;dx\;dt-
\frac{1}{2c} \int_{0}^{t_0-\frac{x_0}{c}} H''(t)(x_0+c(t_0-t))\;dt + \\
& \frac{1}{2c} \int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) \;dx\;dt+ \\
& \frac{x_0}{c} H'\left(t_0-\frac{x_0}{c} \right) - H(t_0)+H\left(t_0-\frac{x_0}{c} \right)
\end{align*}

\begin{align*}
& =\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{c(t_0-t)-x_0} f(t,x)\;dx\;dt + 
\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{x_0+c(t_0-t)} f(t,x)\;dx\;dt + 
\frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) \;dx\;dt+ \\
& \color{blue} \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}}H''(t) (x_0-c(t_0-t))\;dt
-\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}}H''(t) (x_0+c(t_0-t))\;dt\,+\\
& \frac{x_0}{c} H'\left(t_0-\frac{x_0}{c} \right)-H(t_0)+H\left(t_0-\frac{x_0}{c} \right)
\end{align*}

\begin{align*}
\color{blue}
\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} H''(t) (x_0-c(t_0-t))\;dt
-\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} H''(t) (x_0+c(t_0-t))\;dt = \\
\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} H''(t) (-c(t_0-t))\;dt
-\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} H''(t)c(t_0-t)\;dt = \\
-\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} H''(t) c(t_0-t)\;dt
-\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} H''(t)c(t_0-t)\;dt = \\
-\int_{0}^{t_0-\frac{x_0}{c}} H''(t) (t_0-t)\;dt
\end{align*}

Now we use integration by parts, so that 
$$\int_{0}^{t_0-\frac{x_0}{c}}  H''(t)(t_0-t)\,dt =H'(t)(t_0-t)\vert_{0}^{t_0-\frac{x_0}{c}}+\int_{0}^{t_0-\frac{x_0}{c}}  H'(t)dt = \\
\frac{x_0}{c}H'\left(t_0-\frac{x_0}{c}\right)+H\left(t_0-\frac{x_0}{c}\right)-H(0)$$

\begin{align*}
\text{I}& +\text{II}+\text{III} = \\
& \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{c(t_0-t)-x_0} f(t,x)\;dx\;dt + 
\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{x_0+c(t_0-t)} f(t,x)\;dx\;dt + 
\frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) \;dx\;dt \\
& \color{blue}-\frac{x_0}{c}H'\left(t_0-\frac{x_0}{c}\right)-H\left(t_0-\frac{x_0}{c}\right)+H(0)+\\
\color{black}
& \frac{x_0}{c} H'\left(t_0-\frac{x_0}{c} \right)-H(t_0)+H\left(t_0-\frac{x_0}{c} \right)
\end{align*}

\begin{align*}
\text{I}& +\text{II}+\text{III} = \\
& \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{c(t_0-t)-x_0} f(t,x)\;dx\;dt + 
\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{x_0+c(t_0-t)} f(t,x)\;dx\;dt + 
\frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) \;dx\;dt + \\
& H(0) -H(t_0)
\end{align*}

\begin{align*}
& w(t_0,x_0)= \frac{\phi(x_0+ct_0) +\phi(ct_0-x_0)}{2}-H(0) + \\
& \frac{1}{2c} \left[\int_{0}^{ct_0-x_0}\psi(x)\,dx +  \int_{0}^{x_0+ct_0}\psi(x)\,dx \right]-H'(0)t_0 + \\
& \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{c(t_0-t)-x_0} f(t,x)\;dx\;dt + 
\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{x_0+c(t_0-t)} f(t,x)\;dx\;dt + 
\frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) \;dx\;dt + \\
& H(0) -H(t_0)
\end{align*}

\begin{tcolorbox}[colback=white, title=Solution]

So for $ct_0>x_0$, $u(t_0,x_0) =w(t_0,x_0)+H(t_0)$, which means
\begin{align*}
&u(t_0,x_0)= \frac{\phi(x_0+ct_0) +\phi(ct_0-x_0)}{2}+ \\
& \frac{1}{2c} \left[\int_{0}^{ct_0-x_0}\psi(x)\,dx +  \int_{0}^{x_0+ct_0}\psi(x)\,dx \right]-H'(0)t_0 + \\
& \frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{c(t_0-t)-x_0} f(t,x)\;dx\;dt + 
\frac{1}{2c}\int_{0}^{t_0-\frac{x_0}{c}} \int_{0}^{x_0+c(t_0-t)} f(t,x)\;dx\;dt + \\
& \frac{1}{2c}\int_{t_0-\frac{x_0}{c}}^{t_0} \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) \;dx\;dt \\
\end{align*}

While for $ct_0>x_0$, from (\ref{eqn:part1}), we have
$$u(t_0, x_0)=\frac{\phi(x_0+ct_0)+\phi(x_0-ct_0)}{2} +
\frac{1}{2c} \int_{x_0-ct_0}^{x_0+ct_0}\psi(x) dx +
\frac{1}{2c} \int_{0}^{t_0}  \int_{x_0-c(t_0-t)}^{x_0+c(t_0-t)} f(t,x) dx\,dt$$

```{r, warning=FALSE, fig.height=3, fig.width=5, echo=FALSE}
c=1
t_0=5
x_0=3
x_1=seq(-20, 20, by=0.2)
x_2=seq(-20, x_0, by=0.2)
x_3=seq(x_0, 20, by=0.2)
x_4=seq(0, t_0-c*x_0, by=0.2)
rm(x)

ggplot()+
  geom_line(aes(x=x_2, y=t_0-(1/c)*(x_0-x_2)))+
  geom_line(aes(x=x_3, y=t_0+(1/c)*(x_0-x_3)))+
  geom_polygon(aes(x=c(0, 0, x_0, x_0+c*t_0), y=c(0,t_0-x_0/c, t_0,0)), fill="blue", alpha=0.3)+
  geom_polygon(aes(x=c(0, 0, t_0-c*x_0), y=c(0,t_0-x_0/c, 0 )), fill="red", alpha=0.5)+

  lims(y=c(0, 10), x=c(-5, 10))+
  geom_vline(xintercept=0, linetype="dashed")+
  geom_hline(yintercept = t_0-x_0/c, linetype="dashed")+
  geom_text(aes(x=-0.5, y=0.6, label="I"))+
  geom_text(aes(x=3, y=0.6, label="II"))+
  geom_text(aes(x=3, y=t_0-x_0/c+1, label="III"))+
  annotate("text", label="'('* x[0]* ','* t[0] *')'", x=3, y=5.5, size=3, parse = TRUE)+
  labs(x="x", y="t")+
  theme_light()

```

When $x_0<ct_0$, we're integrating $f$ over the domain of dependence (the shaded blue region, including the red triangle) but then adding $f$ integrated over the red triangle. When $x_0<ct_0$, we just integrate $f$ over the blue triangle.

```{r, warning=FALSE, echo=FALSE, fig.height=3, fig.width=5}
x_0=6
x_1=seq(-20, 20, by=0.2)
x_2=seq(-20, x_0, by=0.2)
x_3=seq(x_0, 20, by=0.2)

ggplot()+
  geom_line(aes(x=x_2, y=t_0-(1/c)*(x_0-x_2)))+
  geom_line(aes(x=x_3, y=t_0+(1/c)*(x_0-x_3)))+
  geom_polygon(aes(x=c(x_0-c*t_0,  x_0+c*t_0, x_0) , y=c(0,0, t_0)), fill="blue", alpha=0.3)+
 lims(y=c(0, 10), x=c(-5, 15))+
  geom_vline(xintercept=0, linetype="dashed")+
  geom_hline(yintercept = t_0-x_0/c, linetype="dashed")+
  annotate("text", label="'('* x[0]* ','* t[0] *')'", x=x_0, y=5.5, size=3, parse = TRUE)+
  labs(x="x", y="t")+
  theme_light()
```

@@

Let's assume that we can factor this wave equation with $c=c(x)$ as we could with constant $c$:
$$\left( \partial_{t}+c(x)\partial_{x}\right) \left(\partial_{t}-c(x)\partial_{x}\right)u=0$$
Now, let's distribute:
$$\left( \partial_{t}+c(x)\partial_{x}\right) \left(u_{t}-c(x)u_{x}\right)=0$$
$$u_{tt}-c(x)u_{xt} +c(x)u_{xt}-c(x)\left(c'(x)u_x+c(x)u_{xx}\right) =0$$
$$u_{tt}-c(x)^2u_{xx}- c(x)c'(x)u_x=0$$
But this is only equivalent to our original equation if $c'(x)=0$ for all $x$, i.e., if $c$ is constant.

We can repeat the process for $c=c(t)$:
$$\left( \partial_{t}+c(t)\partial_{x}\right) \left(\partial_{t}-c(t)\partial_{x}\right)u=0$$
$$\left( \partial_{t}+c(t)\partial_{x}\right) \left(u_{t}-c(t)u_{x}\right)=0$$
$$u_{tt}-(c'(t)u_{x}+c(t)u_{xt}) +c(t)u_{xt}-c(t)^2u_{xx} =0$$
$$u_{tt}-c(t)^2u_{xx}- c'(t)u_x=0$$

Which, as before, is only equivalent to our original equation if $c'(t)=0$ for all $t$, i.e., if $c$ is constant.

### 2024-02-17 PDEs/math245_old.Rmd
hw2 pde neumann

>>>>

## HW 2 Problem 6

Does the D’Alembert method work for the wave equation $u_{tt} = c(x)^2u_{xx}$? What about $u_{tt} = c(t)^2u_{xx}$? Why?

Let's try the factorization method if $c=c(x)$:
$$\left( \partial_{t}+c(x)\partial_{x}\right) \left(\partial_{t}-c(x)\partial_{x}\right)u=0$$
Set $\xi=x+c(x)t$ and $\eta=x-c(x)t$.

By the chain rule,

$$\frac{\partial}{\partial x} = \frac{\partial}{\partial \xi} \frac{\partial \xi}{\partial x}+\frac{\partial}{\partial\eta}\frac{\partial\eta}{\partial x}$$

$$=\partial_{\xi}(1+c')+\partial_{\eta}\,(1-c')$$

$$\frac{\partial}{\partial t} = \frac{\partial}{\partial \xi} \frac{\partial \xi}{\partial t}+\frac{\partial }{\partial\eta}\frac{\partial\eta}{\partial t}$$
$$=c\partial_{\xi}-c\partial_{\eta}$$
When we plug these back into
$$\left( \partial_{t}+c\partial_{x}\right) \left(\partial_{t}-c\partial_{x}\right)u=0$$
we get

$$ \left(c\partial_{\xi}-c\partial_{\eta}+c\partial_{\xi}(1+c')+c\partial_{\eta}\,(1-c')\right)
\left(c\partial_{\xi}-c\partial_{\eta} -c\partial_{\xi}(1+c')-c\partial_{\eta}\,(1-c')\right)u=0$$

When $c$ is constant, $c'=0$ and this simplifies to $-4c^2 u_{\xi\eta}=0$, allowing us to integrate to find $u(x,t)=F(x+ct)+G(x-ct)$. But we cannot do the same simplification when we have the $c'$ terms, meaning we cannot solve the wave equation by the same method when the wave speed is not constant.

Similarly,

Let's try the factorization method if $c=c(t)$:
$$\left( \partial_{t}+c(t)\partial_{x}\right) \left(\partial_{t}-c(t)\partial_{x}\right)u=0$$
Set $\xi=x+c(t)t$ and $\eta=x-c(t)t$.

By the chain rule,

$$\frac{\partial}{\partial x} = \frac{\partial}{\partial \xi} \frac{\partial \xi}{\partial x}+\frac{\partial}{\partial\eta}\frac{\partial\eta}{\partial x}=
\partial_{\xi}+\partial_{\eta}$$

$$\frac{\partial}{\partial t} = \frac{\partial}{\partial \xi} \frac{\partial \xi}{\partial t}+\frac{\partial }{\partial\eta}\frac{\partial\eta}{\partial t}=\partial_{\xi}(c't+c)-\partial_{\eta}(c't+c)
$$

When we plug these back into
$$\left( \partial_{t}+c\partial_{x}\right) \left(\partial_{t}-c\partial_{x}\right)u=0$$
we get

$$\left( \partial_{\xi}(c't+c)-\partial_{\eta}(c't+c)+c\partial_{\xi}+c\partial_{\eta} \right)
\left(\partial_{\xi}(c't+c)-\partial_{\eta}(c't+c)-c\partial_{\xi}-c\partial_{\eta}\right)u=0$$
Where again we cannot cancel terms to simplify our equation as we could with constant $c$.

<!-- paste0("(", "x[0]","','","t[0]", ")")) -->
